---
title: "Balanced multi-factor ANOVA"
author: "Sammi Johnson"
date: "2026-01-10"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)
```


## 1 Generating factor levels for manual data input

```{r}
yld <- c ( 69,91,84,76,  66,56,58,60,  8,4,18,10, 44,24,30,22, 
           118,96,80,106, 132,88,101,119)
var <- gl ( n=2, k=12, length=24, labels=c("a","b") )
cli <- gl ( n=3, k=4,  length=24, labels=c("I","II","III") )
rep <- gl ( n=4, k=1,  length=24, labels=c("r1","r2","r3","r4"))
d2 <-data.frame(var,cli,rep,yld)
```


## 2 Two-factor ANOVA: Climate experiment

```{r}
library("GAD" )
cl <- read.table ( "data/v14-u31-01.csv", header=T, sep=";", dec=",",
                  stringsAsFactors  = T)
str(cl)
```

Stripcharts:
```{r}
library("GAD" )
cl <- read.table ( "data/v14-u31-01.csv", header=T, sep=";", dec=",",
                  stringsAsFactors  = T)
str(cl)
```

Climate and variety are fixed factors:
```{r}
cl$C  <- as.fixed (cl$climate)
cl$V  <- as.fixed (cl$variety)
```


Note the difference:
```{r}
class(cl$climate)
class(cl$C)
```


Linear model:
```{r}
class(cl$climate)
class(cl$C)
```


Interaction plots:
```{r}
with ( cl, interaction.plot ( climate, variety, yield, type="b" ))
with ( cl, interaction.plot ( variety, climate, yield, type="b" ))
```


Tables of means and effects:
```{r}
model.tables ( aov(m.1), "means" )
model.tables ( aov(m.1), "effects" )
```


MSE and DFE from the ANOVA table:
```{r}
mse <- gad(m.1)$anova$'Mean Sq'[4]
dfe <- gad(m.1)$anova$'Df'[4]
```


How do you know where these values come from?
```{r}
str(gad(m.1))
gad(m.1)$anova$'Mean Sq'      # column "Mean SQ"
gad(m.1)$anova$'Mean Sq'[4]   # 4th element of column "Mean Sq"
```


### SED, LSD, HSD

```{r}
alpha <- 0.05
n.v   <- 12 ; k.v  <- 2
n.c   <-  8 ; k.c  <- 3
n.cv  <-  4 ; k.cv <- 6
sed.c  <- sqrt ( 2*mse / n.c )  
sed.v  <- sqrt ( 2*mse / n.v )  
sed.cv <- sqrt ( 2*mse / n.cv ) 

t.val <- qt( 1-alpha/2,dfe )
lsd.c  <- t.val * sed.c   
lsd.v  <- t.val * sed.v   
lsd.cv <- t.val * sed.cv  

(hsd.v  <- qtukey( 1-alpha,k.v ,dfe ) * sed.v / sqrt(2) )
(hsd.c  <- qtukey( 1-alpha,k.c ,dfe ) * sed.c / sqrt(2) )
(hsd.cv <- qtukey( 1-alpha,k.cv ,dfe ) * sed.cv/ sqrt(2) )
```


Different values for climate, variety and the combinations:
```{r}
sed.c 
sed.v 
sed.cv

lsd.c 
lsd.v 
lsd.cv

hsd.c 
hsd.v 
hsd.cv
```


## 3 Hierarchical ANOVA: Soils experiment

Data input
```{r}
soils <- read.table ( "data/v14-u31-02a.csv", header=T, sep=";", dec=",",
                     stringsAsFactors = T)
str(soils)
```


Stripchart:
```{r}
stripchart ( yield ~ village:region,
            vertical=T, col="blue",
            data=soils )
```


Region is fixed, village is random:
```{r}
soils$R  <- as.fixed (soils$region)
soils$V  <- as.random(soils$village)
m <- lm ( yield ~ R + R:V, data=soils )
gad (m)
estimates (m)
```


### Means and effects

```{r}
model.tables(aov(m),"means")
model.tables(aov(m),"effects")
```


### LSD

```{r}
va <- data.frame(gad(m)$anova)
dfe <- va[2,1]
mse <- va[2,3]

alpha <- 0.05
b <- 2 ; n <- 4
sed <- sqrt(2 * mse / ( b * n)) 
lsd <- sed * qt(1 - alpha/2,dfe ) 

sed
lsd
model.tables(aov(m), "means" , cterms = "R")  # means only for the regions
```


Analysis as a single-factor ANOVA:
```{r}
str(soils)

soils$RV <- paste("r", soils$R, soils$V, sep = "")
soils$RV <- factor(soils$RV)
levels(soils$RV)
soils$RV <- ordered(soils$RV,levels =
                      c("rIv1","rIv2","rIIv1","rIIv2","rIIIv1","rIIIv2"))
str(soils)

soils$V  <- as.fixed(soils$village)
m.4 <- lm (yield ~ RV, data = soils)
gad(m.4)
```


Treatment means and HSD:
```{r}
model.tables(aov(m.4),"means")

va <- data.frame(gad(m.4)$anova)
(dfe <- va[2,1])
(mse <- va[2,3])

alpha <- 0.05
n <- 4 
a <- 6
sem <- sqrt(mse/(n))
sed <- sqrt(2) * sem 
hsd <- qtukey( 1-alpha,a,dfe ) * sem 

model.tables(aov(m.4),"means")
sed
hsd
```


Display pairwise differences:
```{r}
attach(soils)
m <- tapply(soils$yield,soils$RV,mean)
detach(soils)
source("utils.R")
comp.lsmeans(m,hsd)
```


Function:
```{r}
comp.lsmeans <- function ( lsm, lsd, description.length=4,
                           description.precision=2)
{
  sorted <- sort(lsm,decreasing=T)
  l <- length(sorted)
  cmp  <- NULL
  grps <- 0
  for (i in 1:l) {
    tmp <- matrix ('',ncol=1,nrow=l)
    for (j in i:l) 
      tmp[j] <- if (abs(sorted[i]-sorted[j]) < lsd) 1+grps else ""
    if (i==1) 
    { add <- 1 } 
    else 
    { add <- 0 
    for (j in i:l) 
    { if ( (tmp[j,1]!="") && (cmp[j,grps]=="")) add <- 1 }
    }
    if (add) 
    {colnames(tmp)<-names(sorted[i]) 
    cmp<-cbind(cmp,tmp); grps=grps+1;}
  }
  rownames(cmp) <- paste (format(names(sorted),digits=description.length),
                          round(sorted,description.precision))
  cmp
}
```


## 4 Mixed model ANOVA: Rapeseed experiment

Data input
```{r}
raps <- read.table ( "data/v14-u31-06.csv", header=T, sep=";", dec=",",
                    stringsAsFactors = T)
str(raps)
```


Genotype and environment are fixed, replication is random
```{r}
raps$G  <- as.fixed(raps$geno)
raps$E  <- as.fixed(raps$env)
raps$R  <- as.random(raps$rep)
m <- lm(hgt ~ G*E + E:R, data=raps)
gad(m)
estimates(m)
```


Treatment means for genotypes:
```{r}
model.tables(aov(m), "means", cterms="G")
```


Estimate standard error and dfe:
```{r}
library("lme4")
library("emmeans")
```


```{r}
m2 <- lmer(hgt ~ G + E + G:E + (1|E:R), data = raps)
sem <- data.frame(emmeans(m2,"G"))$SE[1]
dfe <- data.frame(emmeans(m2,"G"))$df[1]
```


### LSD:
```{r}
alpha <- 0.10
sed <- sqrt(2) * sem
lsd <- sed * qt( 1-alpha/2,dfe )
```


Treatment means for GxE interaction:
```{r}
model.tables(aov(m),"means",cterms="G:E")
```


Estimate standard error and dfe:
```{r}
sem <- data.frame(emmeans(m2,"E", by="G"))$SE[1]
```


LSD:
```{r}
alpha <- 0.10
sed <- sqrt(2) * sem
lsd <- sed * qt( 1-alpha/2,dfe )
```


alternative:
```{r}
snk.test ( m, anova.tab=gad(m),
           term ='G', among ='G', within ='E')
snk.test ( m, anova.tab=gad(m),
           term ='G:E', among ='G', within ='E')
```


## 5 Cleaning up

```{r}
rm(list = ls())
```

            
