---
title: "Introduction to R"
author: "Sammi Johnson"
date: "2026-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 2 Setting the working directory 

`Ctrl+Shift+h`

## 3 Using R scripts

- Add the path to your working directory:
- Check if the working directory is set correctly:
- Send command from script to console:



## 4 Comments in R scripts

- You can add comments to scripts by putting a # (hash
- at the beginning of the comment.
- Everything after the hash will be printed as text
- in the console.)


This will be evaluated:

```{r}
1 + 1
```

`# This is just a comment:`

```{r}
5 + 5
```


Code and comments can be combined:

```{r}
1 + 1      # add two numbers
```


## 5 Variables


## 6 Data types


## 7 Data structures


### 7.1 Vectors

Calculation with vectors:



Alphanumeric vectors:



Accessing elements of a vector:


### 7.2 Matrices

For comparison:

Accessing elements of a matrix:

Vectors and matrices contain either numbers or character elements:


### 7.3 Data frames

```
fieldtrial <- data.frame(
  plot         = c( 1,  2,  3,   4,   5,   6,  7,   8,  9,  10,   11,  12),
  variety      = c(                                                      ) ,
  fertilizer   = c(                                                      ),
  rep          = c( 1,  2,  3,   1,   2,   3,  1,   2,  3,   1,    2,   3),
  yield        = c(                                                       ))
```


## 8 Saving and loading data files


## 9 Reading in data from Excel or OpenOffice files

Read from csv file:

```{r}
data = read.csv2("./data/fieldtrial2.csv")
```


## 10 Cleaning up

```{r}
rm(list = ls())
```

